---
title: "Live Session Week 9"
author: "Jeffrey Yau and Devesh Tiwari"
date: "October 26, 2016"
output: html_document
---

## Agenda: 
  1. Time Series Analysis (Continue) - Concept Check (30 minutes total: 15 minutes break-out room)

  2. Time Series Analysis (Continue) - Empirical Exercises (30 minutes total: 15 minutes break-out room)


## 0. Topics Covered This Week:

  1. Notion, Measure, and Estimatino of Dependency and Stationarity

  2. Stationary and non-Stationary Time Series Processes

  3. Random Walk and Autoregressive Process

  4. Time Series Smoothing Technique


## 1. Concepts Check:

Discuss the following concepts:

    i. Define a stochastic process
    
    # A collection of random variables indexed by time.  A time sequence where variables are subject to random variation.
    
    ii. Define a time series
    
    # A series of data points indexed in time order, often taken in equally spaced points in time.
    
    iii. What are some of the most important characteristics we study in a time series so that we can "model" it?
    
    Mean, Variance, Autocorrelation, sampling interval, duration of experiment
    
    iv. Discuss the mean and variance functions and how the similarities and differences from those we studied in classical linear model
    
    Mean depends on time.  Expectaton is over ensemble generated by the stochastic process.   Variance depends on time.  Expectaton is over ensemble generated by the stochastic process.  
    
    v. Define strict and weak statonarity
    
    Strict: any consecutive n sereies of points has same mean and variance
    Weak: any consecutive n series of points has same mean and autocorrelation
    
    vi. How do we measure dependency in a time series?
    
    Autocorrelation
    
    vii. What is the main purpose of time series smoothing?
    
    identify trend and seasonal components.
    
    viii. Suppose your boss give you the monthly sales in two top-performing departments of the company last 5 years. You look at them and find that they both have increasing trends. Your boss ask you to (1) give him the mean of each series and (2) the correlation of the two series. Based on what you learn in the last two lectures, do you think the mean and the correlation are meaningful?
    
    Mean is not meaningful because increasing trend.  Might be useful for comparison.
    Correlation not useful because of both have increasing trend.


## 2. Empirical Exercises:

For each of the given series:

    i. Load the given series into R
    
    ii. Conduct exploratory time series data analysis:
        a. Plot the time series plot (or t-plot) of the series (remember, the graph needs to have a title, well-labeled, and well-chosen tick marks on the axes)
        
      
      
        b. Plot a histogram of the series
```{r}
ts.plot(series1,series2,series3)
```

        c. Provide summary statistics
        
        
        
        d. Construct the correlogram
    
    iii. Make an educated guess the potential stochastic process from which the series could be generated? Please provide your rationale.
```{r}

```



